---
title: "git_phyloseq_RDS"
author: "Thomas Amble Karlsen"
date: "2026-02-08"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F)
```

```{r theme, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Downloads the `rmdformats` package with the document format unless it has been installed previously
```

---

**This file creates an RDS-file containing the phyloseq objects used for downstream analyses**

---

## Loading libraries

```{r}
library(tidyverse)
library(readxl)
library(phyloseq)
library(microViz)
```


## Reading required files

Reading and wrangling required files.

```{r}
# Metadata:
metadata <- read_excel("metadata_reinsdyr.xlsx")

# Read counts:
bracken_counts <- read_tsv("taxprofiler_taxpasta/bracken_PlusPFP_14.07.2025.tsv") %>% 
  filter(!is.na(taxonomy_id)) %>% 
  rename_with(~ str_remove(.x, "_PlusPFP_14.07.2025.bracken"), -taxonomy_id) %>%
  rename_with(~ str_remove(.x, "^\\d+-"), -taxonomy_id)

# Taxonomy lineages:
bracken_lineages <- read_tsv("taxprofiler_taxpasta/bracken_lineages.tsv") %>% 
  rename(taxonomy_id = taxid) %>% 
  filter(!is.na(taxonomy_id))
```

## Creating objects required for phyloseq

Converting tibbles to objects suitable for phyloseq.

```{r}
# Creating count matrix
OTUs_ps <- bracken_counts %>% 
  column_to_rownames("taxonomy_id") %>% 
  as.matrix()

# Creating taxa matrix
taxa_ps <- bracken_lineages %>% 
  filter(taxonomy_id %in% rownames(OTUs_ps)) %>% 
  column_to_rownames("taxonomy_id") %>% 
  as.matrix()

# Creating sample table
samples_ps <- metadata %>% 
  mutate(id = str_replace_all(id, " ", "")) %>% 
  column_to_rownames("id") %>% 
  as.data.frame()
```

## Creating phyloseq object and fixing

Creating phyloseq object and filtering out unknown taxa and human DNA.

```{r}
ps <- phyloseq(otu_table(OTUs_ps, taxa_are_rows = TRUE),
         tax_table(taxa_ps),
         sample_data(samples_ps)) %>% 
  ps_mutate(id_animal = factor(id_animal))

# tax_fix
ps_fix <- ps %>% 
  tax_fix(min_length = 4,
          unknowns = c(""),
          sep = " ", anon_unique = TRUE,
          suffix_rank = "classified") %>% 
  subset_taxa(species != "Homo sapiens") %>%  # removing human DNA
  ps_mutate(season = factor(season),
            location = factor(location))
```

## Subsetting and rarefaction

Creating two subsets of the fixed phyloseq object `ps_fix`, one without control samples (`main_ps`) and one with only control samples (`control_ps`). Additionally creating a rarefied phyloseq object (`rarefied_ps`) using `phyloseq::rarefy_even_depth` with all samples being rarefied to 9727577 reads, the lowest number of reads (R1 & R2) across all samples. Seed 123 used for reproducibility, as `rarefy_even_depth` is random.

```{r, results='hide'}
# subset without control samples
main_ps <- ps_fix %>% 
  subset_samples(!sample_names(ps_fix) %in% c("Microbiotacontrol", "Spike-incontrol"))

# subset with only control samples
control_ps <- ps_fix %>% 
  subset_samples(sample_names(ps_fix) %in% c("Microbiotacontrol", "Spike-incontrol"))

# amount of mapped reads per sample (excluding control samples)
sample_sums(main_ps) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "sample") %>% 
  rename("mapped_reads" = ".") %>% 
  arrange(desc(mapped_reads))

# setting seed for reproducibility
set.seed(123)

# creating a rarefied phyloseq object with depth equal to the sample with the lowest depth
rarefied_ps <- main_ps %>% 
  rarefy_even_depth(sample.size = min(sample_sums(main_ps)))
```

## Creating RDS file

Creating and RDS file containing objects `main_ps`, `control_ps`, `rarefied_ps` and `metadata` for use in other scripts.

```{r}
list_ps <- list(main_ps = main_ps,
                control_ps = control_ps,
                rarefied_ps = rarefied_ps,
                metadata = metadata)

saveRDS(list_ps, file = "rds/phyloseq.rds")
```


