---
title: "ANCOM-BC2"
author: "Thomas Amble Karlsen"
date: "2026-02-12"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F)
```

```{r theme, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Downloads the `rmdformats` package with the document format unless it has been installed previously
```

## Loading libraries

```{r}
library(tidyverse) # general usage (wrangling, plotting...)
library(phyloseq) # exploring the microbiome
library(microViz) # wrangling and visualization of phyloseq data
library(ANCOMBC) # differential abundance analysis
library(ggh4x) # plotting
library(patchwork) # plotting
library(kableExtra) # formatting kable tables
library(tidytext) # column reordering
```

## Loading data

```{r}
readRDS("/Users/thomasamble/Downloads/Master/R_master/rds/phyloseq.rds") %>% 
  list2env(envir = .GlobalEnv)
```

## Running ANCOM-BC2

First filtering out taxa with relative abundances below 0.1%, and subsetting to only bacterial taxa. Then converting back to count data, the required input for ANCOM-BC2. Then running ANCOM-BC2 with season and location as fixed effects, at species level, with a prevalence cutoff at 10%, meaning the species have to be present in at least 10% of the samples. ANCOM-BC2 was run twice, once grouped by season and once grouped by location. First running grouped by season:

```{r}
metadata <- metadata %>% 
  mutate(season = factor(season))

holm_season <- main_ps %>% 
  tax_transform("compositional") %>% 
  tax_filter(min_sample_abundance = 0.001, tax_level = "unique",
             use_counts = FALSE, prev_detection_threshold = 0) %>% 
  ps_get(counts = TRUE) %>% 
  subset_taxa(top_rank == "Bacteria") %>% 
  ancombc2(fix_formula = "season + location",
           group = "season",
           p_adj_method = "holm",
           tax_level = "species",
           global = TRUE,
           pairwise = TRUE,
           alpha = 0.05,
           prv_cut = 0.10)
```

```{r}
clean_season <- holm_season$res_pair %>% 
  rename(species = taxon) %>% 
  left_join(main_ps@tax_table %>% as.data.frame(),
              by = "species") %>% 
  pivot_longer(cols = matches("^(lfc|se|q|passed_ss)_"),
               names_to = c(".value", "contrast"),
               names_pattern = "(lfc|se|q|passed_ss)_(.*)") %>%
  select(species, phylum, contrast, lfc, se, q, passed_ss) %>% 
  filter(q < 0.05, passed_ss == TRUE, abs(lfc) > 1) %>% 
  mutate(contrast = case_match(contrast,
                               "seasonSummer" ~ "Summer vs Autumn (adjusted for location)",
                               "seasonWinter" ~ "Winter vs Autumn (adjusted for location)",
                               "seasonWinter_seasonSummer" ~ "Winter vs Summer (adjusted for location)"),
         species = reorder_within(species, lfc, contrast))

plot_season <- clean_season %>% 
  ggplot(aes(x = lfc, y = reorder(species, lfc),
             xmin = lfc - se, xmax = lfc + se, fill = phylum)) +
  geom_errorbar(width = 0.5) +
  geom_point(shape = 21, size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray") +
  facet_wrap(facets = vars(contrast), scales = "free_y", space = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("Actinomycetota" = "#A6CEE3", 
                                "Bacillota" = "#1F78B4",
                                "Pseudomonadota" = "#B2DF8A",
                                "Bacteroidota" = "#33A02C",
                                "Verrucomicrobiota" = "#6A3D9A",
                                "Chloroflexota" = "#9F008A")) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none",
        axis.text.y = element_text(face = "italic")) +
  labs(x = "log2 fold change", y = "",
       title = "A) ANCOM-BC2 grouped by season")
```

Now running grouped by location:

```{r}
holm_location <- main_ps %>% 
  tax_transform("compositional") %>% 
  tax_filter(min_sample_abundance = 0.001, tax_level = "unique",
             use_counts = FALSE, prev_detection_threshold = 0) %>% 
  ps_get(counts = TRUE) %>% 
  subset_taxa(top_rank == "Bacteria") %>% 
  ancombc2(fix_formula = "season + location",
           group = "location",
           p_adj_method = "holm",
           tax_level = "species",
           global = TRUE,
           pairwise = TRUE,
           alpha = 0.05,
           prv_cut = 0.10)

clean_location <- holm_location$res_pair %>% 
  rename(species = taxon) %>% 
  left_join(main_ps@tax_table %>% as.data.frame(),
              by = "species") %>% 
  pivot_longer(cols = matches("^(lfc|se|q|passed_ss)_"),
               names_to = c(".value", "contrast"),
               names_pattern = "(lfc|se|q|passed_ss)_(.*)") %>%
  select(species, phylum, contrast, lfc, se, q, passed_ss) %>% 
  filter(q < 0.05, passed_ss == TRUE, abs(lfc) > 1) %>% 
  mutate(contrast = case_match(contrast,
                               "locationRøros" ~ "Røros vs Mo i Rana (adjusted for season)",
                               "locationTana_locationRøros" ~ "Tana vs Røros (adjusted for season)",
                               "locationTana" ~ "Tana vs Mo i Rana (adjusted for season"),
         species = reorder_within(species, lfc, contrast))

plot_location <- clean_location %>% 
  ggplot(aes(x = lfc, y = reorder(species, lfc),
             xmin = lfc - se, xmax = lfc + se, fill = phylum)) +
  geom_errorbar(width = 0.3) +
  geom_point(shape = 21, size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray") +
  facet_wrap(facets = vars(contrast), scales = "free_y", space = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("Actinomycetota" = "#A6CEE3", 
                                "Bacillota" = "#1F78B4",
                                "Pseudomonadota" = "#B2DF8A",
                                "Bacteroidota" = "#33A02C",
                                "Verrucomicrobiota" = "#6A3D9A",
                                "Chloroflexota" = "#9F008A")) +
  theme_bw(base_size = 14) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(x = "log2 fold change", y = "",
       title = "B) ANCOM-BC2 grouped by location")
```

Plotting both ANCOM-BC2 plots together with `patchwork`.

```{r}
plot_season + plot_location +
  plot_layout(guides = "collect")

ggsave("plots/ancombc2_holm.png", width = 15, height = 10)
```

Creating supplementary tables with exact values of ANCOM-BC2 metrics. 

```{r}
holm_season$res_pair %>% 
  rename(species = taxon) %>% 
  left_join(main_ps@tax_table %>% as.data.frame(),
              by = "species") %>% 
  pivot_longer(cols = matches("^(lfc|se|q|passed_ss)_"),
               names_to = c(".value", "contrast"),
               names_pattern = "(lfc|se|q|passed_ss)_(.*)") %>%
  select(species, phylum, contrast, lfc, se, q, passed_ss) %>% 
  filter(q < 0.05, passed_ss == TRUE, abs(lfc) > 1) %>% 
  mutate(contrast = case_match(contrast,
                               "seasonSummer" ~ "Summer vs Autumn",
                               "seasonWinter" ~ "Winter vs Autumn",
                               "seasonWinter_seasonSummer" ~ "Winter vs Summer")) %>%
  arrange(contrast, desc(lfc)) %>% 
  print(n = 41) %>% 
  DT::datatable(options = list(scrollY = TRUE))


holm_location$res_pair %>% 
  rename(species = taxon) %>% 
  left_join(main_ps@tax_table %>% as.data.frame(),
              by = "species") %>% 
  pivot_longer(cols = matches("^(lfc|se|q|passed_ss)_"),
               names_to = c(".value", "contrast"),
               names_pattern = "(lfc|se|q|passed_ss)_(.*)") %>%
  select(species, phylum, contrast, lfc, se, q, passed_ss) %>% 
  filter(q < 0.05, passed_ss == TRUE, abs(lfc) > 1) %>% 
  mutate(contrast = case_match(contrast,
                               "locationRøros" ~ "Røros vs Mo i Rana",
                               "locationTana_locationRøros" ~ "Tana vs Røros",
                               "locationTana" ~ "Tana vs Mo i Rana")) %>% 
  arrange(contrast, (desc(lfc))) %>% 
  print(n = 22) %>% 
  DT::datatable(options = list(scrollY = TRUE))
```


