---
title: "bash scripts"
author: "Thomas Amble Karlsen"
date: "2026-02-08"
output: 
  rmdformats::downcute:
    self_contained: true  
    thumbnails: true  
    lightbox: true  
    gallery: false  
    highlight: tango  
    code_folding: show
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      eval = F,
                      message = F,
                      warning = F)
```

```{r theme, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Downloads the `rmdformats` package with the document format unless it has been installed previously
```

---

**This file contains all scripts used in the Sigma2 high-throughput computing cluster saga. All #SBATCH information is excluded.**

---


## Initial QC

Performing quality control using `FastQC` (version 0.11.9).

```{bash}
input=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/reads
output=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/fastqc/results
files=($(ls ${input}/*.fastq.gz))
input_file=${files[$SLURM_ARRAY_TASK_ID]}


# Activate modules and envs
module load Miniconda3/23.10.0-1
source ${EBROOTMINICONDA3}/bin/activate
conda activate /cluster/projects/nn9305k/src/miniconda/envs/fastqc

# Run
fastqc --outdir $output $input_file
```


## Host genome indexing with Bowtie2

Indexing the genome of the host organism, Rangifer tarandus, using `bowtie2-build` (version 2.5.1).

```{bash}
# Paths
genome_zip=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/bowtie2_indexing/GCA_949782905.1_mRanTar1.h1.1_genomic.fna.gz
prefix=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/bowtie2_indexing/GCA_949782905.1_mRanTar1.h1.1_genomic
genome_unzip="${prefix}.fna"

# Activating modules and environments
module load Bowtie2/2.5.1-GCC-12.3.0

# Unzipping genome file
gunzip -c "$genome_zip" > "$genome_unzip"

# Running bowtie2-build
bowtie2-build --large-index --threads 24 "$genome_unzip" "$prefix"
```

## Taxprofiler

Running the Taxprofiler pipeline from nf-core. Using a minimum read length of 36 bp for QC. Performing run merging to concatenate FASTQ files from different sequencing lanes. Running Kraken2+Bracken, Centrifuge and MetaPhlAn for taxonomic classification.

```{bash}
module load Java/21.0.2


/cluster/projects/nn9305k/bin/nextflow_25.04.7 run /cluster/projects/nn10070k/pipelines/nf-core-taxprofiler_1.2.3/1_2_3/main.nf \
-profile apptainer \
-work-dir $USERWORK/taxprofiler \
--input /cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/taxprofiler/samplesheet.csv \
--databases /cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/taxprofiler/databasesheet.csv \
--outdir /cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/taxprofiler \
--multiqc_title "MultiQC_remotnitor" \
--perform_shortread_qc \
--perform_shortread_complexityfilter \
--shortread_qc_minlength 36 \
--perform_shortread_hostremoval \
--hostremoval_reference /cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/bowtie2_indexing/GCA_949782905.1_mRanTar1.h1.1_genomic.fna \
--shortread_hostremoval_index /cluster/projects/nn10070k/projects/remotnitor/thomas_master/data/bowtie2_indexing \
--perform_runmerging \
--run_centrifuge \
--run_kraken2 \
--run_bracken \
--bracken_save_intermediatekraken2 \
--run_metaphlan \
--run_profile_standardisation \
--run_krona \
-c /cluster/projects/nn10070k/configs/saga_taxprofiler.config \
-resume
```


## TaxonKit & csvtk

This script creates a taxonomy table to supply the Bracken table for integration into a phyloseq object, using TaxonKit and csvtk.

```{bash}
# Stop the script if anything fails
set -e

# Setup:
in_file=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/taxprofiler/taxpasta/bracken_PlusPFP_14.07.2025.tsv
tax_ids=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/taxprofiler/taxpasta/taxids.txt
database=/cluster/projects/nn10070k/databases/kraken2_pluspfp_14.07.2025
out_file=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/taxprofiler/taxpasta/bracken_lineages.tsv

# First: Creating a list of all taxonomy ids in the bracken file
     # cut -f1: extract the first column
     # tail -n +2: remove the header (print everything starting from line 2)
     # grep '^[0-9]\+$': keeps only numeric rows
     # sort -u: sort and deduplicate taxonomy ids

cut -f1 $in_file \
| tail -n +2 \
| grep '^[0-9]\+$' \
| sort -u > $tax_ids

# Then: Creating the table with lineages
     # taxonkit reformat2: builds and reformats a lineage table to include the wanted ranks
          # -f: specifies format, {superkingdom|domain|acellular root} means choose whichever exists, in that specified order
          # -r "": missing ranks are left empty
          # -T: do not fill lower ranks to be the same as the higher ranks
     # csvtk add-header: adds a header
     # csvtk sep: splits the lineage_string column into the different ranks
     # csvtk cut: keeps only the relevant columns

taxonkit reformat2 \
  --data-dir $database \
  -f '{superkingdom|domain|acellular root};{phylum};{class};{order};{family};{genus};{species}' \
  -r '' -T \
  $tax_ids \
| csvtk -t add-header -n taxid,lineage_string \
| csvtk -H -t sep -f lineage_string -s ';' -N 7 -R --merge \
     -n top_rank,phylum,class,order,family,genus,species \
| csvtk -H -t cut -f taxid,top_rank,phylum,class,order,family,genus,species \
> $out_file
```

## Removing reads mapped to several taxa from MetaPhlAn results

Removing deduplicated reads (reads that map to several marker genes) from the MetaPhlAn results.

```{bash}
cd /cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/taxprofiler/metaphlan/mpa_vJan25_CHOCOPhlAnSGB_202503

out_file=/cluster/projects/nn10070k/projects/remotnitor/thomas_master/results/taxprofiler/metaphlan/metaphlan_bowtie2_merged_mapped.tsv

# creating header:
echo -e "sample\tmapped_reads" > "$out_file"

# looping over all .metaphlan.bowtie2out.txt-files
for file in *.metaphlan.bowtie2out.txt; do
    sample=$(basename "$file" .metaphlan.bowtie2out.txt)

    # count number of unique read ids mapped:
        # cut -f1: extracts the first column (read ids)
        # sort -u: removes deduplicates within the first column
        # wc -l: counts how many unique mapped reads are left after deduplication
    mapped=$(cut -f1 | sort -u | wc -l)

    # merging results from each file into a new file
    echo -e "${sample}\t${mapped}" >> "$out_file"
done
```


