---
title: "ANCOM-BC2"
author: "Thomas Amble Karlsen"
date: "2026-02-12"
output: 
  rmdformats::downcute:
    self_contained: true  
    thumbnails: true  
    lightbox: true  
    gallery: false  
    highlight: tango  
    code_folding: show
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F)
```

```{r theme, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Downloads the `rmdformats` package with the document format unless it has been installed previously
```

## Loading libraries

```{r}
library(tidyverse) # general usage (wrangling, plotting...)
library(phyloseq) # exploring the microbiome
library(microViz) # wrangling and visualization of phyloseq data
library(ANCOMBC) # differential abundance analysis
library(ggh4x) # plotting
```

## Loading data

```{r}
readRDS("/Users/thomasamble/Downloads/Master/R_master/rds/phyloseq.rds") %>% 
  list2env(envir = .GlobalEnv)
```

## Running ANCOM-BC2

First filtering out taxa with relative abundances below 0.1%, and subsetting to only bacterial taxa. Then converting back to count data, the required input for ANCOM-BC2. Then running ANCOM-BC2 with season and location as fixed effects, at species level, with a prevalence cutoff at 10%, meaning the species have to be present in at least 10% of the samples. ANCOM-BC2 was run twice, once using p-value adjustment method Holm-Bonferroni (holm), and once using Benjamini-Hochberg (bh).

```{r}
ancombc_holm <- main_ps %>% 
  tax_transform("compositional") %>% 
  tax_filter(min_sample_abundance = 0.001, tax_level = "unique",
             use_counts = FALSE, prev_detection_threshold = 0) %>% 
  ps_get(counts = TRUE) %>% 
  subset_taxa(top_rank == "Bacteria") %>% 
  ancombc2(fix_formula = "season + location",
           tax_level = "species",
           p_adj_method = "holm",
           alpha = 0.05,
           prv_cut = 0.10)

ancombc_bh <- main_ps %>% 
  tax_transform("compositional") %>% 
  tax_filter(min_sample_abundance = 0.001, tax_level = "unique",
             use_counts = FALSE, prev_detection_threshold = 0) %>% 
  ps_get(counts = TRUE) %>% 
  subset_taxa(top_rank == "Bacteria") %>% 
  ancombc2(fix_formula = "season + location",
           tax_level = "species",
           p_adj_method = "BH",
           alpha = 0.05,
           prv_cut = 0.10)
```


## Plotting

Plotting significantly differentially abundant species for both p-value adjustment methods, and coloring by phylum.

```{r}
# Holm
p_holm <- ancombc_holm$res %>% rename(species = taxon) %>% 
  left_join(main_ps@tax_table %>% as.data.frame() %>% rownames_to_column("tax_id"),
              by = "species") %>% 
  pivot_longer(cols = matches("^(lfc|se|q|passed_ss)_"),
               names_to = c(".value", "contrast"),
               names_pattern = "(lfc|se|q|passed_ss)_(.*)") %>% 
  filter(contrast != "(Intercept)", q < 0.05) %>% 
  select(species, phylum, contrast, lfc, se, q, passed_ss) %>% 
  mutate(contrast = case_match(contrast,
                               "seasonSummer" ~ "Summer vs Autumn (ref.)\nadjusted for location",
                               "locationRøros" ~ "Røros vs Mo i Rana (ref.)\nadjusted for season")) %>% 
  ggplot(aes(x = lfc, y = reorder(species, lfc),
             xmin = lfc - se, xmax = lfc + se, fill = phylum)) +
  geom_errorbar(width = 0.05) +
  geom_point(shape = 21, size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray") +
  facet_wrap(facets = vars(contrast), ncol = 1, scales = "free_y") +
  labs(x = "logFoldChange", y = "",
       title = "A) Holm-Bonferroni-adjusted") +
  scale_fill_manual(values = c("Actinomycetota" = "#A6CEE3", 
                                "Bacillota" = "#1F78B4",
                                "Pseudomonadota" = "#B2DF8A")) +
  theme_bw() +
  theme(axis.text.y = element_text(face = "italic"),
        legend.position = "none",
        plot.title = element_text(hjust = 1))


# BH
p_bh <- ancombc_bh$res %>% rename(species = taxon) %>% 
  left_join(main_ps@tax_table %>% as.data.frame() %>% rownames_to_column("tax_id"),
              by = "species") %>% 
  pivot_longer(cols = matches("^(lfc|se|q|passed_ss)_"),
               names_to = c(".value", "contrast"),
               names_pattern = "(lfc|se|q|passed_ss)_(.*)") %>% 
  filter(contrast != "(Intercept)", q < 0.05) %>% 
  select(species, phylum, contrast, lfc, se, q, passed_ss) %>% 
  mutate(contrast = case_match(contrast,
                               "seasonSummer" ~ "Summer vs Autumn (ref.)\nadjusted for location",
                               "locationRøros" ~ "Røros vs Mo i Rana (ref.)\nadjusted for season")) %>% 
  ggplot(aes(x = lfc, y = reorder(species, lfc),
             xmin = lfc - se, xmax = lfc + se, fill = phylum)) +
  geom_errorbar(width = 0.35) +
  geom_point(shape = 21, size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray") +
  facet_wrap(facets = vars(contrast), ncol = 1, scales = "free_y") +
  labs(x = "logFoldChange", y = "",
       title = "B) Benjamini-Hochberg-adjusted") +
  scale_fill_manual(values = c("Actinomycetota" = "#A6CEE3", 
                                "Bacillota" = "#1F78B4",
                                "Pseudomonadota" = "#B2DF8A")) +
  theme_bw() +
  theme(axis.text.y = element_text(face = "italic"),
        legend.position = "bottom",
        legend.justification = c(1.25,0),
        plot.title = element_text(hjust = 1)) +
  force_panelsizes(rows = c(1,8))

p_holm + p_bh

ggsave("ancombc2_comp.png")
```









