---
title: "Abundances"
author: "Thomas Amble Karlsen"
date: "2026-02-12"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F)
```

```{r theme, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Downloads the `rmdformats` package with the document format unless it has been installed previously
```

## Loading libraries

```{r}
library(tidyverse) # general usage (wrangling, plotting...)
library(phyloseq) # exploring the microbiome and importing into R from Kraken-BIOM
library(microViz) # visualization of microbiome data
library(patchwork) # plotting
```

## Loading data

```{r, results='hide'}
readRDS("/Users/thomasamble/Downloads/Master/R_master/rds/phyloseq.rds") %>% 
  list2env(envir = .GlobalEnv)
```

## Control samples

Creating a barplot of the top 12 species for the two control samples. Also creating tables of the top 10 most abundant species in the Microbiota control sample. Finally creating a table of the relative abundances of the 10 species present in the *ZymoBIOMICS Microbial Community Standard* for the Microbiota control sample, alongside the expected relative abundances of these species.

```{r}
control_ps %>% 
  comp_barplot(tax_level = "species", n_taxa = 12)

control_ps %>% 
  subset_samples(sample_names(control_ps) == "Microbiotacontrol") %>% 
  tax_transform("compositional") %>% 
  psmelt() %>% 
  select(OTU, Sample, Abundance, species) %>% 
  rename(taxid = OTU) %>% 
  slice(1:10)

# Bacillus spizizenii is a subspecies of Bacillus subtilis

# Numbers and species from ZymoBIOMICS Microbial Community Standard:
zymobiomics_species <- c("Pseudomonas aeruginosa", 
                         "Escherichia coli",
                         "Salmonella enterica", 
                         "Limosilactobacillus fermentum",
                         "Enterococcus faecalis", 
                         "Staphylococcus aureus",
                         "Listeria monocytogenes",
                         "Bacillus subtilis",
                         "Saccharomyces cerevisiae",
                         "Cryptococcus neoformans")

control_ps %>% 
  subset_samples(sample_names(control_ps) == "Microbiotacontrol") %>%
  tax_transform("compositional") %>% 
  psmelt() %>% 
  select(OTU, Sample, Abundance, species) %>% 
  rename(taxid = OTU,
         `Observed relative abundance` = Abundance,
         Species = species) %>% 
  filter(Species %in% zymobiomics_species) %>% 
  mutate(Species = factor(Species, levels = zymobiomics_species),
         `Expected relative abundance` = case_when(Species %in% c("Pseudomonas aeruginosa",
                                                                  "Escherichia coli",
                                                                  "Salmonella enterica",
                                                                  "Limosilactobacillus fermentum",
                                                                  "Enterococcus faecalis",
                                                                  "Staphylococcus aureus",
                                                                  "Listeria monocytogenes",
                                                                  "Bacillus subtilis") ~ 12 / 100,
                                                   Species %in% c("Saccharomyces cerevisiae",
                                                                  "Cryptococcus neoformans") ~ 2 / 100),
         Difference = `Observed relative abundance` - `Expected relative abundance`,
         log2FC = log2(`Observed relative abundance` / `Expected relative abundance`)) %>% 
  select(Species, `Observed relative abundance`, `Expected relative abundance`, Difference, log2FC) %>% 
  arrange(Species)
```

## Wrangling and plotting

Reading in the kraken-inspect file from the PlusPFP database, extracting and plotting the relative abundances of taxa Bacteria, Eukaryota, Archaea and Viruses. Also plotting relative abundances of the same taxa for all samples in the rarefied phyloseq object. Additionally plotting the relative abundances of the top 12 phyla from the rarefied phyloseq object. Combining into one plot. Finally calculating median relative abundances for both the top-level taxonomy rank and the phylum rank on the rarefied data.

```{r}
inspect <- read_table("inspect.txt",
                      comment = "#",
                      col_names = c("percent","kmers","taxa","rank","taxid","name"))

toptax_levels <- c("Bacteria", "Eukaryota", "Archaea", "Viruses")

plot_db <- inspect %>% 
  filter(taxid %in% c(2, 2157, 2759, 10239)) %>% 
  mutate(abundance = percent / 100,
         name = factor(name, levels = toptax_levels),
         stack_order = fct_reorder(name, abundance)) %>% 
  ggplot(aes(x = "", y = abundance, fill = name, group = stack_order)) +
  geom_col(width = 0.2) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set3",
                    breaks = c("Bacteria", "Eukaryota", "Archaea", "Viruses")) +
  labs(title = "A) Top-level taxonomy (DB)",
       x = "",
       y = "Relative abundance") +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90),,
        plot.title = element_text(hjust = 0.50),
        axis.ticks.y = element_line()) +
  guides(fill = guide_legend(title = "Top-level taxonomy"))
  

plot_toptax <- rarefied_ps %>% 
  comp_barplot(tax_level = "top_rank") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.75, hjust = 1, margin = margin(t = -9)),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        plot.title = element_text(hjust = 0.50)) +
        #legend.position = "inside",
        #legend.key.size = unit(0.9, "lines")) +
  labs(y = "",
       title = "B) Top-level taxonomy") +
  scale_fill_brewer(palette = "Set3",
                    breaks = toptax_levels)


plot_phyla <- rarefied_ps %>% 
  comp_barplot(tax_level = "phylum",
               n_taxa = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.75, hjust = 1, margin = margin(t = -9)),
        axis.ticks.x = element_blank(),
        plot.title = element_text(hjust = 0.50)) +
  labs(title = "C) Phylum",
       y = "") +
  guides(fill = guide_legend(title = "Phylum"))
  

plot_db + plot_toptax + plot_phyla +
  plot_layout(widths = c(0.40, 1.2, 1.35), guides = "collect") &
  theme(legend.title = element_text(size = 13),
        legend.text  = element_text(size = 9),
        legend.key.size = unit(0.9, "lines"),
        legend.key.height = unit(0.9, "lines"),
        legend.key.width  = unit(0.9, "lines"),
        legend.key = element_rect(color = "black", linewidth = 0.3),
        legend.background = element_blank())

ggsave("abundance_comp.png", width = 12)


# top top_rank
rarefied_ps %>% 
  tax_agg("top_rank") %>% 
  tax_transform("compositional") %>%
  psmelt() %>% 
  group_by(OTU) %>% 
  summarise(median_RA = median(Abundance)*100) %>% 
  arrange(desc(median_RA))


# top phyla
rarefied_ps %>% 
  tax_agg("phylum") %>% 
  tax_transform("compositional") %>%
  psmelt() %>% 
  group_by(OTU) %>% 
  summarise(median_RA = median(Abundance)*100) %>% 
  arrange(desc(median_RA))
```

## Statistics

Aggregating the rarefied phyloseq object to phylum rank, and transforming to relative abundances ("compositional"). Then filtering to only keep bacterial taxa, and filtering out phyla with a mean relative abundance below 1% across all samples. Then fitting an ANOVA with sample-level phylum relative abundance as response variable, and an interaction term with season \* location \* phylum (main effects and interactions) as explanatory variables. Then performing a TukeyHSD-test on the ANOVA, and extracting contrasts where the compared groups belong to the same phylum, and have p-values below 0.05.

```{r}
phylum_data <- rarefied_ps %>%
  tax_agg("phylum") %>%
  tax_transform("compositional") %>% 
  psmelt() %>% 
  #filter(Sample != "WKR24-79") %>% 
  filter(top_rank == "Bacteria") %>% 
  group_by(phylum) %>%
  mutate(mean_abundance = mean(Abundance)) %>% 
  ungroup() %>%
  filter(mean_abundance > 0.01)

# ANOVA
## Note: Relative abundance is not independent, increase one phylums abundance -> decrease others (violates assumptions)
a <- aov(Abundance ~ season * location * phylum,
    data = phylum_data)

# Tukey HSD
Tukey <- TukeyHSD(a, conf.level = 0.95)

same_phylum <- Tukey$`season:location:phylum` %>%
  as.data.frame() %>% 
  rownames_to_column() %>%
  separate_wider_delim(rowname,
                       delim = "-",
                       names = c("obs_1", "obs_2")) %>% 
  separate_wider_delim(obs_1,
                       delim = ":",
                       names = c("season_1", "location_1", "phylum_1")) %>% 
  separate_wider_delim(obs_2,
                       delim = ":",
                       names = c("season_2", "location_2", "phylum_2")) %>% 
  filter(phylum_1 == phylum_2,
         `p adj` < 0.05)

same_phylum
```
